Standard Candle Cosmology With Incomplete Spectroscopic Classification

### 2015-10-16 doc/article.tex
Keep on writing article

\section{Standard Candle}
To introduce basic model elements, this Section considers a
simplified view of SNe~Ia
as transients that are bolometric standard candles with an intrinsic random
dispersion
at peak brightness, and assumes data with negligible measurement uncertainties.
For convenience,
the transients
in this Section are  referred to as SNe~Ia.
The added complexity of correlations between absolute magnitude
and multi-band light curves, and measurement uncertainties
are discussed in
\S\ref{snIamodel:sec}.

@@

\section{SN~Ia Model}
\label{snIamodel:sec}
\subsection{Base Model}
SNe~Ia are not perfect standard candles but rather are calibrated candles whose
peak brightnesses are a
function of their internal properties discerned from their colors and light-curve evolution.
Measurement uncertainty, particularly from photometric flux
calibration and galaxy photometric redshifts, can play an important role in the statistical
reach of experiments.  This Section contains a model that expands upon the one
presented in \S\ref{snIamodel:sec} to have the fidelity needed for the photometric
supernova analysis of the Dark Energy Survey and the Large Synoptic Survey Telescope.

The likelihood is written as
\begin{equation}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G |  \Omega_M, w, \theta_r,\alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}}, \theta_{Ia}, \theta_{non-Ia}, Z, g, g_N),
\label{likelihood2:eqn}
\end{equation}
whose relations are sketched in the probabilistic graphical model in Figure~\ref{pgm:fig}.

@@

The new measurement that differs from  \S\ref{model:sec} is
\begin{itemize}
\item $\mathit{ADU}$: the measured counts for each photometric observation takes
the place of flux;
\end{itemize}
The new model parameters are
\begin{itemize}
\item $\theta_{Ia}$, $\theta_{non-Ia}$: the parameters for the internal diversity within SNe~Ia and
non-Ia's respectively;
\item $Z$: the parameters that describe the transmission functions, i.e.\ photometric calibration.
\end{itemize}
New latent model parameters for each transient:
\begin{itemize}
\item $\theta_{Ia,i}$, $\theta_{non-Ia,i}$: the underlying parameters that describe the transient;
\item $g$: the parameters that describe the candidate host and neighbor galaxies including redshift
is a more general replacement for $z$.
\end{itemize}
The new fixed parameters are
\begin{itemize}
\item{$z$}: the cosmological transient redshift given the host galaxy;
\item{$L$}: the luminosity of the transient with specified parameters;
\item{$f$, $f_g$}: the flux of the transient and foreground hosts;
\item{$\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$}: the expected counts of the
transient and foreground galaxies.
\end{itemize}

\subsubsection{Likelihood for the obs set data}
The likelihood of data from the {\it obs} set is
\begin{equation}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_S, \theta_G|  \Omega_M, w, \theta_r,\alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}}, \theta_{Ia}, \theta_{non-Ia}, Z, g, g_N).
\end{equation}
Again assuming that spectroscopy gives an unambiguous redshift, type and now adding
host-galaxy properties 
\begin{align}
p(\theta_S, \theta_G|g) &= \delta(\{\theta_S, \theta_G\} -g)
\end{align}
then
\begin{multline}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_S, \theta_G|  \ldots) = \\
p(\mathit{ADU} | T=T_S, z=z_s, g=\{\theta_S, \theta_G\},  \dots) P(T_S| z= z_S , g=\{\theta_S, \theta_G\}, \ldots).
\label{obs3:eqn}
\end{multline}

The first term is the likelihood for the light curve.  The dependence of the light curve depends
on transient properties, cosmology, calibration, and the experiment.  For convenience. the
construction of the likelihood is broken up into several steps, with intermediary points
at the determination of the time and wavelength-dependent luminosity $L$,
the flux $f$, the expected counts $\overline{\mathit{ADU}}$ of the transient and background, and
finally the observed light curve $\mathit{ADU}$.  Figure~\ref{pgm:fig} shows how these
parameters are related to the intrinsic parameters and to each other.

The pdf of the luminosity $L$ can be expressed as follows:
\begin{multline}
p(L|  \theta_r,\alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}}, \theta_{Ia}, \theta_{non-Ia}, g)  \\
= \sum_T \int d\theta_{Ia}\, d\theta_{non-Ia}\, P(L, T, \theta_{Ia} \theta_{non-Ia}|\ldots) \\
= \sum_T \int d\theta_{Ia}\, d\theta_{non-Ia}\, P(L| T, \theta_{Ia} \theta_{non-Ia} \ldots) p(T,
\theta_{Ia}\, \theta_{non-Ia} | \ldots)
\end{multline}

\subsubsection{Likelihood for the mis set data}

%The likelihood for the model is
%\begin{multline}
%p(\mathit{ADU}, {{T}}_S,{{z}}_S,
%{{\theta}}_S, \theta_G, \tau, Z | g, \theta_\mu, \theta_r, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia}, \phi) = \\
%p(\theta_G|g)p(Z|\phi)p(\mathit{ADU}, {{T}}_S,{{z}}_S,
%{{\theta}}_S, \tau | g, \theta_\mu, \theta_r, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia}, \phi).
%\label{likelihood:eqn}
%\end{multline}

The simulation of data from the model presented in
and their analysis is implemented in Python and are available
on GitHub\footnote{{git@github.com:AlexGKim/abc.git}}.

>>>>

\section{Standard Candle}
\label{toy:sec}
To introduce basic model elements, this Section considers a
simplified view of SNe~Ia
as transients that are bolometric standard candles with an intrinsic random
dispersion
at peak brightness, and assumes data with negligible measurement uncertainties.
For convenience,
the transients
in this Section are  referred to as SNe~Ia.
The added complexity of correlations between absolute magnitude
and multi-band light curves, and measurement uncertainties
are discussed in
\S\ref{snIamodel:sec}.

@@

\section{DES SN Survey Model}
\label{snIamodel:sec}
\subsection{Model}
SNe~Ia are not perfect standard candles but rather are calibrated candles whose
peak brightnesses are a
function of their internal properties discerned from their colors and light-curve evolution.
Measurement uncertainty, particularly from photometric flux
calibration and galaxy photometric redshifts, can play an important role in the statistical
reach of experiments.  This Section contains a model that expands upon the one
presented in \S\ref{snIamodel:sec} to have the fidelity needed for the photometric
supernova analysis of the Dark Energy Survey and the Large Synoptic Survey Telescope.

The notation for the likelihood presents the data and fundamental model parameters
\begin{equation}
p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G |  \Omega_M, w, \theta_r, \alpha_{Ia}, \beta_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\beta_{\mathit{non-Ia}}, \sigma_{\mathit{non-Ia}},  Z, g, g_N, f_g),
\label{likelihood2:eqn}
\end{equation}
whose relations are sketched in the probabilistic graphical model in Figure~\ref{pgm:fig}.

@@

The new measurement that differs from  \S\ref{model:sec} is
\begin{itemize}
\item $\mathit{ADU}$: the measured counts for each photometric observation.
This takes
the place of flux as calibration is now considered.
\end{itemize}
The new model parameters are
\begin{itemize}
\item $Z$: the parameters that describe the transmission functions, i.e.\ photometric calibration.
\item $\beta_{Ia}$, $\beta_{non-Ia}$: the parameters for the internal diversity within SNe~Ia and
non-Ia's respectively, e.g.\ SALT2's  $\alpha$ and $\beta$;
\item $f_g$: the surface brightness of the background light at the location of the transient
needed for to extract the photometry of the point-source transient.
\end{itemize}
New latent model parameters for each transient:
\begin{itemize}
\item $\theta_{Ia}$, $\theta_{non-Ia}$: the transient parameters that describe individual
transients, e.g.\ SALT2's $x_1$, $c$;
\item $g$: the parameters that describe the candidate host and neighbor galaxies,
to be more expansive than the previous $z$ parameter.
\end{itemize}
The new fixed parameters are
\begin{itemize}
\item{$f$}: the flux of the transient;
\item{$\overline{\mathit{ADU}}_X$}: the expected counts of the transient and background light.
\end{itemize}

The relationships between the parameters for the underlying population are now described.  What follows is an initial proposal
based on current supernova analyses, but many different models need to be
considered as the understanding of transients improve
and for probing systematics by comparing the
performance of competing models.

\begin{itemize}
\item Intrinsic transient parameters are dependent on their distributions and the host galaxy.
The SNe~Ia parameters are drawn as
\begin{equation}
\theta_{Ia} | \sigma_{Ia}, g \sim \mathcal{N}(0, \sigma_{Ia}),
\end{equation}
where $\theta_{Ia} =\{x_1, c, \delta_M\}$, the SALT2 light-curve shape and color parameters.
A more complex model can be based on non-Normal distributions
and/or including SN properties directly tied to characteristics (e.g.\ mass, star formation
rate) of the host galaxy.

The non-Ia parameters are drawn as
\begin{equation}
\theta_{non-Ia} | \sigma_{non-Ia}, g  \sim \mathcal{N}(0, \sigma_{non-Ia}),
\end{equation}
where the parameter is the realization of an intrinsic magnitude dispersion.
A more complex model can include multiple non-Ia populations each with its own
intrinsic parameters.
\item Spectroscopically observed transient parameters depend on the intrinsic transient
parameters and type.  $\theta_{Si}$ may be ignored in a zero'th order analysis that
does not relate spectral features with SN~Ia absolute magnitudes.   If spectral
features are considered, the simple model
\begin{equation}
\theta_{Si} | \theta_{Ia}, \theta_{non-Ia}, T = \delta(\theta_{Si}-\theta_T)
\end{equation}
assumes an unambiguous mapping between supernova parameters and spectral signal.
\item The type is drawn from the host-dependent relative rate between SNe~Ia and non-Ia's 
\begin{equation}
T | \theta_r, g \sim \rm{Bernoulli}(\pi(g,\theta_r)),
\end{equation}
where $\pi$ is a model for the relative rates of the different types as a
function of galaxy parameters and redshift.  A reasonable choice for $\pi$ needs to be
selected, any volunteers?
\item Supernova type is assumed to be unambiguously encoded in the
spectrum:
\begin{equation}
T_S | T  = \delta(T_S - T).
\end{equation}
\item The observed redshift is taken to be the host galaxy redshift
\begin{equation}
z_S = g_z,
\end{equation}
where $g_z$ is the element in $g$ that corresponds to redshift.
Added sophistication includes redshift
measurement uncertainty and random peculiar velocities. The study of coherent
flow requires added model complexity.
\item The underlying SED of transients are modeled.  For SNe~Ia the SALT2 model
is an option, although new models that separate dust and intrinsic color or
identify new SN~Ia parameters are expected in the near future.  Absolute magnitudes
that depend on host-galaxy properties need to be considered. 
For SNe~Ia $T=1$
\begin{equation}
L(t,\lambda; \alpha_{Ia}, \beta_{Ia}, \theta_{Ia}, g)  =10^{-\left(\frac{\alpha x_1 + \beta c}{2.5}\right)}L_{SALT2}(t-t_0; \alpha_{Ia}, x_1, c, \delta_m) \gamma(g, \beta) ,
\end{equation}
where the first two terms is the SALT2 model with a realized error snake,
and $\gamma(g, \beta)$ a step function with free parameters in $\beta$ that
describe the step size.

For non-Ia's $T=0$ a set of light curve templates or a model can be used.
An initial candidate is the treatment of non-Ia's implemented in PSNID.
\item The theoretical distance modulus
\begin{equation}
\mu(g_z; \Omega_M, w)
\end{equation}
is the predicted distance modulus of a flat dark energy cosmology.
\item The flux incident on Earth is the redshifted and time-dilated SED scaled
by the distance modulus.
\begin{equation}
f(t,\lambda) = 10^{-\mu/2.5} L\left(t/(1+g_z),\lambda/(1+g_z)\right)
\end{equation}
\item The expected light curve in counts depends on the flux and throughput of the
optical chain.  A first model encapsulates flux calibration through a zeropoint
\begin{equation}
\overline{\mathit{ADU}}(f, Z) = 10^{Z/2.5} \int  f(\lambda) T_0(\lambda)d\lambda,
\end{equation}
where $T_0$ is a fiducial transmission function for the appropriate filter.  A more sophisticated
analysis can have a transmission function specific to each observation.
\item The measured surface brightness of the host galaxy depends on the underlying
surface brightness convolved with the point spread function and integrated
over the transmission
\begin{equation}
\overline{\mathit{ADU}}_g =    \int (f_g(\lambda, RA, Dec) \otimes PSF) T(\lambda;Z)d\lambda.
\end{equation}
\item The measurement $\mathit{ADU}$ is the superposition of transient and background
light $\overline{\mathit{ADU}} + \overline{\mathit{ADU}}_g$.
\end{itemize}

\subsection{Likelihood}

\subsubsection{Measurements}
In the following discussion,
spectroscopic and galaxy measurements are assumed
to provide prefect determinations of transient type and parameters, and
galaxy redshifts and parameters.  The working group is striving to get spectroscopic
redshifts of all potential hosts.

The measured counts depend on the expected counts and measurement uncertainty
$C_P$ as a Poisson (or considering the high sky background Normal) distribution
\begin{equation}
\mathit{ADU} \sim \mathcal{N} (\overline{\mathit{ADU}}+\overline{\mathit{ADU}}_g ,C_P).
\label{ADU:eqn}
\end{equation}

Practically, the photometric extraction is handled by independent software, which calculates
the pdf of $\mathit{ADU}_{\mathit{SN}}$, the realized contribution of supernova light,
replacing this and the previous two bullet points.  Dillon is providing this.
For the rest of the discussion, $\mathit{ADU}_{\mathit{SN}}$ is treated a measurement
directly connected to the expected mean signal of the supernova
$\overline{\mathit{ADU}}$.

\subsubsection{Likelihood}

The above describes the model for the underlying distribution.  Needed
is the distribution for the population that enter the analysis; detection and spectroscopic
follow-up have parameter-dependent selection efficiencies.  For the analysis sample
imposing slightly more stringent criteria at the expense
of losing a few captured transients may be more convenient than reconstructing
the live-time decision making process.

%The likelihood for the analysis sample is related to the underlying likelihood through
%\begin{multline}
%p(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G |  \Omega_M, w, \theta_r, \alpha_{Ia}, \beta_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\beta_{\mathit{non-Ia}}, \sigma_{\mathit{non-Ia}},  Z, g, g_N, f_g)\\
%= \frac{\epsilon(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G) p_U(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | \ldots)}
%{\sum_{T_S} \int d\mathit{ADU}\, dz_S\, d\theta_S\, d\theta_G\, \epsilon(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G) p_U(\mathit{ADU}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | \ldots)},
%\end{multline} 
%where $p_U$ is the underlying likelihood  and $\epsilon$ is the efficiency for inclusion
%in the sample (more precisely subsample as spectroscopic
%selection should produce different efficiencies for the {\it obs} and {\it mis} sets.)

%The efficiencies for photometric and spectroscopic selection are taken to be independent.
%The efficiency  to pass photometric selection is denoted as $\epsilon_P$.
The acceptance criteria based on photometry has not yet defined; the following is a sketch
of how it would be considered for a criterion of $\ge N$ photometric points
above a signal-to-noise threshold (translating to a distinct limiting signal $ADU_0$
for each measurement of each transient) given $M$ available measurements. 
The fail/pass status of each of the measurements is denoted as
$b$ of 0/1, and there are $\sum_{k=N}^M \binom{M}{k}$ 
distinct $b$'s that allow acceptance of the light curve.

\subsubsection{Likelihood for the {\it obs} Set}
The likelihood for the {\it obs} set is
\begin{multline}
p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G |  \Omega_M, w, \theta_r, \alpha_{Ia}, \beta_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\beta_{\mathit{non-Ia}}, \sigma_{\mathit{non-Ia}},  Z, g, g_N, f_g)\\
= \sum_i \int_L dL\, p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | b_i, L, \ldots)
P(b_{i} |L, \ldots) P(L|\ldots) ,
\end{multline} 
where $L$ are the latent parameters.
The first term is the likelihood for the data given the condition for the  $b_i$ configuration
of threshold pass/fails  and all parameters.  The second term and
third terms are the probabilities of achieving the configuration
and latent parameters
respectively.

The likelihood for one of the $b$ configurations is derived from the likelihood without the configuration condition
\begin{equation}
p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | b_i, L, \ldots)
%=\frac{
% \epsilon(\mathit{ADU}_{SN}; b_i)
%  p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | L, \ldots)}
%  {\sum_{T_S} \int d\mathit{ADU}_{SN}\, dz_S\, d\theta_S\, d\theta_G\, 
% \epsilon(\mathit{ADU}_{SN}; b_i)
%  p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | L, \ldots)}\\
  =  \frac{ \prod_j
 \epsilon(\mathit{ADU}_{SN,j}; b_{ij})
  p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | L, \ldots)}
  {\prod_j P(b_{ij}{ |L,\dots)}} ,
\label{obsconf:eqn}
\end{equation}
where $j$ indexes photometric points and
$\epsilon$ is the efficiency of acceptance of that point from the underlying distribution.
For flux thresholds, the efficiency is
\begin{equation}
\epsilon(\mathit{ADU}_{SN,j}; b_{ij})= \mathcal{H}\left((-1)^{b_{ij}+1}(ADU_{SN,j}-ADU_{0,j})\right),
\end{equation}
where $\mathcal{H}$ is the Heaviside step function; when $b_{ij}=1$ the step in the efficiency
is up and for $b_{ij}=0$ the step is down.

The likelihood without the configuration condition is
\begin{multline}
p(\mathit{ADU}_{SN}, {{T}}_S,{{z}}_S, \theta_{S}, \theta_G | L, \ldots) \\
= \delta(z_S - g_z) \delta(\theta_S-\theta_T) \delta(\theta_G -g) \delta(T_S-T)  \mathcal{N}
\left(\mathit{ADU}_{SN} -\overline{\mathit{ADU}}_j(L ,\ldots), \sigma \right),
\end{multline}
where $\mathcal{N}$ is an $M$-dimensional Normal distribution
and $\sigma$ is the photometric uncertainty.

The probability for failing (passing) the selection criterion for a single point is
\begin{equation}
P(b_{ij} | L, \ldots) = b_{ij} + (-1)^{b_{ij}}\int_{-\infty}^{ADU_0} d\mathit{ADU}_{SN,i}  \mathcal{N}\left(\mathit{ADU}_{SN} - \overline{\mathit{ADU}}_j(L ,\ldots), \sigma_T \right),
\end{equation}
the (complementary) cumulative distribution function for a normal distribution
with variable
$\mathit{ADU}_0$.

\subsubsection{Likelihood for the {\it mis} Set}
The likelihood for the {\it mis} set does not have to account for spectroscopic data
\begin{multline}
p(\mathit{ADU}_{SN}, \theta_G |  \Omega_M, w, \theta_r, \alpha_{Ia}, \beta_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\beta_{\mathit{non-Ia}}, \sigma_{\mathit{non-Ia}},  Z, g, g_N, f_g)\\
= \sum_i \int_L dL\, p(\mathit{ADU}_{SN}, \theta_G | b_i, L, \ldots)
P(b_{i} |L, \ldots) P(L|\ldots) ,
\end{multline} 
\begin{equation}
p(\mathit{ADU}_{SN}, \theta_G | b_i, L, \ldots)
  =  \frac{ \prod_j
 \epsilon(\mathit{ADU}_{SN,j}; b_{ij})
  p(\mathit{ADU}_{SN}, \theta_G | L, \ldots)}
  {\prod_j P(b_{ij}{ |L,\dots)}}.
\label{misconf:eqn}
\end{equation}

Suppose Ravi provides for each transient a set of potential host galaxies
each with an associated probability.  For the case of two potential hosts,
the likelihood without the configuration condition is
\begin{multline}
p(\mathit{ADU}_{SN}, \theta_{G}| L, \ldots) = \\
 \left(\pi_{host} p(z-\theta_{G1})\delta(z_N-\theta_{G2}) +
	(1-\pi_{host}) \delta(z-\theta_{G2})\delta(z_N-\theta_{G1})
\right) \mathcal{N}
\left(\mathit{ADU}_{SN} -\overline{\mathit{ADU}}_j(L ,\ldots), \sigma \right);
\end{multline}
this is trivially generalized for only one or more than two potential hosts.
FIX FOR PHOTO-Z.

All other terms are as for the {\it obs} Set case.

\subsubsection{Spectroscopic Selection}
The efficiency for a successful spectroscopic observation $\epsilon_S$ has
to be determined.  There are volunteers.

\subsubsection{Measurement Uncertainties}
The measurements $\mathit{ADU}$

Measurement uncertainties.
$Z$ has a prior for calibration.

\section{Misc}
The simulation of data from the model presented in
and their analysis is implemented in Python and are available
on GitHub\footnote{{git@github.com:AlexGKim/abc.git}}.
